Scheduling Musical Events as Procedures with Scheme for Max

### 2021-06-23 paper-1/paper-1.tex
added citations

%% edited, needs citations and a diagram
Max (also known as Max/MSP) is a programming environment for creating interactive 
music and multi-media programs through a visual programming language accessible 
to non-programmers.  Programs, or ``patches" in the Max nomenclature, are created by placing
visual boxes representing Max objects on a canvas and connecting them visually with
``patch-cords", a paradigm similar to that of modular synthesizers and familiar to many musicians.
First created in the mid 1980's by Miller Puckette while at IRCAM 
(Institut de recherche et coordination acoustique/musique), 
Max is now developed and sold by the San Francisco software company Cycling '74,
and is widely used in both academic and commercial music contexts as well
as in multi-media installations. A rich library of Max objects exists, 
both provided with Max and available as open-source 
extensions, enabling users to rapidly create interactive systems with gestural 
input from physical sources such as on screen widgets, physical electronic instruments,
and custom hardware communicating over serial networks.

@@

One powerful feature of Max is its ability to be programmed
while the engine is playing music. Patches can be altered 
without necessarily interrupting patch activity (depending on the design
of the program), and this can
even be performed live, an activity known as ``live-coding", which has given
birth to an emerging musical sub-culture of live programming performances
with tools such as Max, SuperCollider, Csound, and others. 
%% TODO citation%%
This approach to music making overlaps with the related discipline of 
algorithmic music, in which programmatic algorithms are used
not just for affecting musical parameters such as volume, pitch, and timbre, 
but also for the generation of musical content such as melodies
and rhythms.

@@

Scheme for Max (a.k.a. S4M) is an open-source external developed by the 
author that embeds the s7 Scheme interpreter, a Scheme dialect created by
Bill Schottstaedt at the  Center for Computer Research in Music and Acoustics
(CCRMA) at Stanford University. Originally developed
from TinyScheme, s7 is a Scheme dialect designed for use in computer music
platforms and is used in various music programs such as the Snd editor and the
Common Music algorithmic composition platform. 
%% TODO citation for s7 here
Similar in functionality to the built
in \textbf{js} object, the \textbf{s4m} object enables the user to program Max in Scheme,
and provides Scheme functions to interact with the Max engine and environment through
a foreign function interface implemented in C using the Max C SDK and API.

@@

Max is a visual programming environment for interactive multi-media, 
used widely in music academia as well as in commercial music circles through "Max for Live", 
a version of Max embedded in the Ableton Live digital audio workstation. 
Max patches are created by placing visual boxes on a canvas 
and connecting them graphically with ``patch cords", where a box may be
any of the Max object types installed on the user's system. A box placed on the patcher
results in the instantiation of an object in memory from a prototypical class for the object, with
text fields typed in the visual patching box used as constructor arguments. 
Thus a Max patch consists of a collection of instantiated objects that send messages to each other
in a directed graph, producing a data-flow execution model whereby a message from a source 
object triggers execution in one or more 
receiving objects, who may in turn send on messages to other similarly connected objects.

Patch activity, in the form of messages moving through the graph, can be initiated by 
various forms of real-time input, such as keyboard and mouse events, connected electronic instruments, and networking 
events, as well as by scheduled events through Max objects such as the \textbf{metronome,} which sends
out messages at regular time intervals. Messages are 
stored internally as lists of Max ``atoms", which may be symbols, integers, or floating point numbers. 
There also exists a Max object for visually displaying and altering messages called the \textbf{message-box},
which allows messages to be typed directly in to the object in the visual patcher, and then sent by clicking the box.  
Execution follows a depth-first and right-to-left order, enabling the programmer to deterministically 
control the execution flow with the visual layout of the patch cords. (i.e., A source object sending 
messages out to multiple receiving sub-graphs results in the right hand message path completing 
execution to the bottom before moving left, rather than spawning two concurrent threads of execution.) When 
patching, messages can be inspected by sending them to a \textbf{print} object (which prints to the Max console),
to a \textbf{message-box} object (which will update its visual display of the message) 
or though a built in visual debugger using a feature Max calls ``probing”.

@@

%% section done except citations

@@

% cite CSound %

@@

While Max is a commercial, closed-source product, it includes a software development kit for 
extending Max by writing a Max “external”. An external is a compiled plug-in that defines the  
prototype class (data structure and methods) used to create new objects in the patcher.  
Externals are developed in C or C++ in an object-oriented manner, the C API using data 
structures and pointers to simulate class-based programming and the more recent C++ API using C++ classes.
A typical external will implement a class that provides some object state for instantiated
objects along with constructor and destructor functions, and  
methods for sending and receiving Max messages through the object’s inlets and outlets.
The SDK and API also provide a rich body of functions for interacting with the overall Max
environment, thus these methods may also trigger side-effects through these facilities.

As externals are compiled plug-ins, they do not need to be distributed as source-code, and
are most typically made available as binaries for Windows and OSX. Extending Max through externals has 
been possible since very early versions of Max, thus thousands of 3rd party externals 
now exist, both open and closed source, of which Scheme For Max is one.

%% cite the object database??
%% done %%

\subsection{Lisp in Computer Music}
%% done, needs finalcopy edit and citation%%

Scheme For Max is far from the first Lisp-based computer music tool, or even
the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
ranging from Common Lisp Music, created by Bill Schottstaedt in the 1980's, to more recent
systems such as Common Music 3, %% cite taube %%
Nyquist, %% cite dannenburg %% 
and Extempore (and its predecessor Impromptu). %% cite impromptu paper
Nor is Scheme for Max unique in providing a Lisp
based extension facility to a commercial music platform - the author unknowingly
programmed in Lisp in the early 1990's while using the Cakewalk sequencer,
which came with an embedded Lisp interpreter in the form of the 
Cakewalk Application Language (CAL).

@@

In contrast with S4M, systems such as Common Music, Nyquist, and Extempore
all run as stand-alone applications, with events originating from them often destined
for other systems for translation into audio. While these systems are, like Max, also capable
of receiving gestural input, they must, in essence, `be their own boss" - they provide
their own scheduler and timing clocks and act as the principal engine from which temporal
events originate. If they are to be used alongside another temporal engine, the two
must be synchronized and events sent back and forth over network connections of some kind.

There have also been previous Max externals created to allow one to use Lisp in Max. 
However, these have been implemented as object that act as proxies to 
external Lisp processes rather than interpreters fully embedded and running in the Max process.
This difference means that in contrast to these offerings, the S4M interpreter has full 
access to the Max C API and can call API functions through foreign function interface
calls, with the results of said calls being exactly the same as if the functions were called in C. 
The ramification of this is that S4M is unique among Max Lisp projects in being
able to operate within the native Max scheduling system, enabling highly accurate
timing and retaining deterministic control flow within a patcher. (i.e., There
is no hidden networking layer with communication to an outside process, which 
makes operations fundamentally asynchronous even if this is hidden from the user visually.)
This provides the user with opportunities to use Scheme for tasks beyond
what is possible with pre-existing options. For example, we are in effect 
able to say ``do this Max-internal thing based on other internal Max state at such a time,
as determined by the Max scheduler". An example of this could be to control
the Max transport mechanism at a certain music time, based on data in a table 
that is used as a shared data store between multiple patcher objects.

\section{implementation of the Scheme for Max external}
% STATUS: in progress

@@

The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
The most serious
is that it is limited to running only in the Max low-priority GUI thread. As this
thread is also used for Max file i/o and graphic redrawing, the result is that latency
is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, and perceives events separated by 20 ms as discrete
events. 
% citeation
Delays of tens of milliseconds are enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the \textbf{js} object is unreliable.

@@

The scheduling system describe does have several limitations worth noting.
The s7 Scheme implementation includes a tracing garbage collector 
that must run occasionally, pausing other processing. 
%% cite bil email
The garbage collector
can, however, be paused and run on demand. Tests on the author's
system, with the garbage collector forced to run every 100ms to prevent
significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
when run in a substantial application (a 16 track sequencer written entirely in S4M
with the computer also generating audio for the tracks with commercial
software synthesizers).  
The audio rendered was recorded and examined
examined in a digital audio workstation to check for timing discrepancies. The
results of these preliminary tests were that timing was unaffected by the 
garbage collector so long as Max was configured with 
a large enough output buffer to allow the collector to run within the 
latency period of this buffer.
Using Max with an output buffer of 256 samples produces an output latency
of approximately 6-8ms (depending on the Max signal vector as well), which
was large enough for the collector to run without issue, but is 
low enough to be acceptable for real time performance. With software synthesis
loads at closer to the maximum possible on the author's computer,
an output buffer of 512 samples (~11 ms output latency) was necessary to prevent
any timing errors. This is encouraging as these buffer sizes are not atypical
for commercial music production on digital audio workstations.

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass. The
size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are designed to compensate for this jitter,
maintaining long term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publically documented or available
in open-source code, but one can assume that it is some variant of the pattern 
whereby scheduled events are aware of when they are supposed to be running
and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the theoretical accurate time of the scheduling event, rather than its
actual (post-jitter) time. The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece - 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.

>>>>

Max (also known as Max/MSP) is a programming environment for creating interactive 
music and multi-media programs through a visual programming language accessible 
to non-programmers. Max was created to make it possible for users
to go beyond the limits of commercial music sequencing tools, creating
interactive environments of arbitrary complexity and sophistication.
\cite{Zicarelli2002}
Programs, or ``patches" in the Max nomenclature, are created by placing
visual boxes representing Max objects on a canvas and connecting them visually with
``patch cords", a paradigm similar to that of modular synthesizers and familiar to many musicians.
First created in the mid 1980's by Miller Puckette while at IRCAM 
(Institut de recherche et coordination acoustique/musique), 
Max is now developed and sold by the San Francisco software company Cycling '74,
and is widely used in both academic and commercial music contexts as well
as in multi-media installations. A rich library of Max objects exists, 
both provided with Max and available as open-source 
extensions, enabling users to rapidly create interactive systems with gestural 
input from physical sources such as on screen widgets, physical electronic instruments,
and custom hardware communicating over serial networks.

@@

One powerful feature of Max is its ability to be programmed
while the engine is playing music. Patches can be altered 
without necessarily interrupting patch activity (depending on the design
of the program), and this can
even be performed live, an activity known as ``live-coding", which has
emerged since the early 2000's as its own musical sub-culture of 
live programming performances
with tools such as Max, SuperCollider, Csound, and others. 
\cite{Roberts18}
This approach to music making overlaps with the related discipline of 
algorithmic music, in which programmatic algorithms are used
not just for affecting musical parameters such as volume, pitch, and timbre, 
but also for the generation of musical content such as melodies
and rhythms.

@@

Scheme for Max (a.k.a. S4M) is an open-source external developed by the 
author that embeds the s7 Scheme interpreter, a Scheme dialect created by
Bill Schottstaedt at the  Center for Computer Research in Music and Acoustics
(CCRMA) at Stanford University. Originally developed
from TinyScheme, s7 is a Scheme dialect designed for use in computer music
platforms and is used in various music programs such as the Snd editor and the
Common Music algorithmic composition platform. 
\cite{Schottstaedt2021} 
Similar in functionality to the built
in \textbf{js} object, the \textbf{s4m} object enables the user to program Max in Scheme,
and provides Scheme functions to interact with the Max engine and environment through
a foreign function interface implemented in C using the Max C SDK and API.

@@

Max is a visual programming environment for interactive multi-media, 
used widely in music academia as well as in commercial music circles through Max for Live, 
a version of Max embedded in the Ableton Live digital audio workstation. 
Max patches are created by placing visual boxes on a canvas 
and connecting them graphically with ``patch cords", where a box may be
any of the Max object types installed on the user's system. A box placed on the patcher
results in the instantiation of an object in memory from a prototypical class for the object, with
text fields typed in the visual patching box used as constructor arguments. 
Thus a Max patch consists of a collection of instantiated objects that send messages to each other
in a directed graph, producing a data-flow execution model whereby a message from a source 
object triggers execution in one or more 
receiving objects, who may in turn send on messages to other similarly connected objects.

Patch activity, in the form of messages moving through the graph, can be initiated by 
various forms of real-time input, such as keyboard and mouse events, connected electronic instruments, and networking 
events, as well as by scheduled events through Max objects such as the \textbf{metronome,} which sends
out messages at regular time intervals. Messages are 
stored internally as lists of Max ``atoms", which may be symbols, integers, or floating point numbers. 
\cite{Puckette2002}
There also exists a Max object for visually displaying and altering messages called the \textbf{message-box},
which allows messages to be typed directly in to the object in the visual patcher, and then sent by clicking the box.  
Execution follows a depth-first and right-to-left order, enabling the programmer to deterministically 
control the execution flow with the visual layout of the patch cords. (i.e., A source object sending 
messages out to multiple receiving sub-graphs results in the right hand message path completing 
execution to the bottom before moving left, rather than spawning two concurrent threads of execution.) When 
patching, messages can be inspected by sending them to a \textbf{print} object (which prints to the Max console),
to a \textbf{message-box} object (which will update its visual display of the message) 
or though a built in visual debugger using a feature Max calls ``probing”.

@@

@@

% TODO: cite Dannenburg on this point  %

@@

While Max is a commercial, closed-source product, it includes a software development kit for 
extending Max by writing a Max “external”. An external is a compiled plug-in that defines the  
prototype (data structure and methods) used to create new objects in the patcher.  
Externals are developed in C or C++ in an object-oriented manner, the C API using data 
structures and pointers to simulate class-based programming with dynamic binding,
and the more recent C++ API using C++ classes \cite{Zicarelli2002}.
A typical external will implement a class that provides some object state for instantiated
objects along with constructor and destructor functions, and  
methods for sending and receiving Max messages through the object’s inlets and outlets.
Methods are bound dynamically to the object prototypes
The SDK and API also provide a rich body of functions for interacting with the overall Max
environment, thus these methods may also trigger side-effects through these facilities.

As externals are compiled plug-ins, they do not need to be distributed as source-code, and
are most typically made available as binaries for Windows and OSX. Extending Max through externals has 
been possible since very early versions of Max in the late 1980's
thus thousands of 3rd party externals 
now exist \cite{MaxObjects2021}, both open and closed source, of which Scheme For Max is one.

\subsection{Lisp in Computer Music}

Scheme For Max is far from the first Lisp-based computer music tool, or even
the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
ranging from Common Lisp Music, created by Bill Schottstaedt in the late 1980's
\cite{Wang17} and Heinrich Taube's Common Music in 1991 \cite{Taube91},
to Dannenburg's Nyquist in 1997 \cite{Dannenberg97},
and Impromptu (later rewritten as Extempore) \cite{Sorensen2010}. 
Nor is Scheme for Max unique in providing a Lisp
based extension facility to a commercial music platform - the author unknowingly
programmed in Lisp in the early 1990's while using the Cakewalk sequencer,
which came with an embedded Lisp interpreter in the form of the 
Cakewalk Application Language (CAL).

@@

In contrast with S4M, systems such as Common Lisp Music, Common Music, Nyquist, and Extempore
all run as stand-alone applications, with events originating from them either
rendered as audio internally or destined for other systems for rendering to audio. 
While these systems are, like Max, also capable
of receiving gestural input, they must, in essence, ``be their own boss" - they provide
their own scheduler and timing clocks and act as the principal engine from which temporal
events originate. If they are to be used alongside another temporal engine, the two
must be synchronized and events sent back and forth over network connections of some kind.

There have also been previous Max externals created to allow one to use Lisp in Max in various ways.
Brad Garton authored MaxLispJ, an external that provides a Common Lisp interpreter 
(Armed Bear Common Lisp) embedded through the Java runtime that can be used within Max 
via the \textbf{mxj} object \cite{Garton2011}.
MOZ'Lib, by Julien Vincenot, takes another approach, in which the Max external
acts as a proxy to an externally running Steel Bank Common Lisp process,
a similar approach to that used by Cycling '74 for their \textbf{Node for Max} object (for Node.JS/ECMA6), 
thus enabling the outside process to execute long-running work without blocking Max \cite{Vincenot2017}.

In contrast to these, S4M embeds a light-weight Scheme interpreter directly in a Max
C external with no intermediate or external runtime, possible because s7 is implemented entirely
in ANSI C and is statically linked during compilation of S4M. 
This means the \textbf{s4m} object has direct C-level access to the Max C API 
through foreign function interface calls, with the results of said calls being exactly 
the same as if the functions were called in C, save execution time. 
The ramification of this is that S4M is unique among Max Lisp projects in being
able to operate within the native Max scheduling system, enabling highly accurate
timing and retaining deterministic control flow within a patcher. (i.e., There
is no hidden networking layer with communication to an outside process, which 
makes operations fundamentally asynchronous even if this is hidden from the user visually.)
This provides the user with opportunities to use Scheme for tasks beyond
what is possible with pre-existing options. For example, we are in effect 
able to say ``do this Max-internal thing based on other internal Max state at such a time,
as determined by the Max scheduler". An example of this could be to control
the Max transport mechanism at a certain music time, based on data in a table 
that is used as a shared data store between multiple patcher objects.

\section{implementation of the Scheme for Max external}

@@

The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
The most serious is that it is limited to running only in the Max low-priority GUI thread \cite{Cycling74}.
 As this
thread is also used for Max file i/o and graphic redrawing, the result is that latency
is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, and perceives events separated by 20 ms as discrete
events. \
Delays of tens of milliseconds are enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
musical sequencers or algorithmic music engines, the \textbf{js} object is unreliable.

@@

The scheduling system describe does have several limitations worth noting.
The s7 Scheme implementation includes a tracing garbage collector 
that must run occasionally, pausing other processing \cite{Matheussen2020}.
The garbage collector can, however, be paused and run on demand. Tests on the author's
system, with the garbage collector forced to run every 100ms to prevent
significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
when run in a substantial application (a 16 track sequencer written entirely in S4M
with the computer also generating audio for the tracks with commercial
software synthesizers).  
The audio rendered was recorded and examined
examined in a digital audio workstation to check for timing discrepancies. The
results of these preliminary tests were that timing was unaffected by the 
garbage collector so long as Max was configured with 
a large enough output buffer to allow the collector to run within the 
latency period of this buffer.
Using Max with an output buffer of 256 samples produces an output latency
of approximately 6-8ms (depending on the Max signal vector as well), which
was large enough for the collector to run without issue, but is 
low enough to be acceptable for real time performance. With software synthesis
loads at closer to the maximum possible on the author's computer,
an output buffer of 512 samples (~11 ms output latency) was necessary to prevent
any timing errors. This is encouraging as these buffer sizes are not atypical
for commercial music production on digital audio workstations and are widely
considered in acceptable ranges for real-time music systems \cite{Brandt1998}.

In addition to output latency, there is the factor of internal latency.
Max, like many computer music programming environments, generates audio 
in vectors of samples, with
each vector generated from one processing pass, and (normally) control
rate passes conducted also a maximum of once per audio vector pass. The
size of this signal vector thus has a large effect on performance, with larger
vectors requiring significantly less CPU use. As audio rendering of note onsets originating
from control messages (the normal scenario) must thus be aligned with 
signal vector boundaries, this produces delays of note onsets by up to
the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
However, the more recent scheduling facilities in Max, including the 
clock facilities used by S4M, are designed to compensate for this jitter,
maintaining long term temporal accuracy despite note onset delays. The mechanism
by which this is done is unfortunately not publically documented or available
in open-source code, but one can assume that it is some variant of the common pattern 
whereby scheduled events are aware of when they are supposed to be running
(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
events will be shifted to compensate for the jitter, based 
on the logical time of the scheduling event, rather than its
actual (post-jitter) real time \cite{Anderson1986}.
The author also conducted personal tests of 
this compensation, comparing both high and low signal vector sizes to produce
variable delays in note onsets, and it was determined
that so long as the overall output latency is again large enough to allow
the GC to run in time, self-scheduling events in a long piece do stay
accurate in that jitter does not accumulate over the piece - 
note onsets stay accurate to within the delay of one signal vector of samples,
regardless of the length of the piece of music.
